---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
---

```{r}
library(knitr)
library(dplyr)
library(ggplot2)
```


---
Loading and preprocessing the data
---


```{r, echo=TRUE}
activity <- read.csv("activity.csv")
head(activity)
summary(activity)
activitydata <- activity[ with (activity, { !(is.na(steps)) } ), ]
```

Calculate and report the mean and median of the total number of steps taken per day
```{r, echo=TRUE}

steps_by_interval <- aggregate(steps ~ interval, activitydata, mean)

# create a time series plot 
plot(steps_by_interval$interval, steps_by_interval$steps, type='l', 
     main="Average number of steps over all days", xlab="Interval", 
     ylab="Average number of steps")
```

#What is mean total number of steps taken per day?


```{r, echo=TRUE}
group_by_day <- group_by(activitydata, date)
total_steps_by_day <- summarise(group_by_day, total = sum(steps))
hist(total_steps_by_day$total, main="Histogram of total number of steps per day", xlab="Total number of steps in a day")
mean(total_steps_by_day$total, na.rm=TRUE)
median(total_steps_by_day$total, na.rm=TRUE)
```
The mean and median are respectively 10766 and 10765 steps taken per day. 

#What is the average daily activity pattern?
```{r, echo=TRUE}
# preprocessing data for plot
steps_by_interval <- aggregate(steps ~ interval, activitydata, mean)
# create a time series plot 
plot(steps_by_interval$interval, steps_by_interval$steps, type='l', 
     main="Average number of steps over all days", xlab="Interval", 
     ylab="Average number of steps")
```

The plot shows a peek around the 800th interval.

```{r, echo=TRUE}
steps_by_interval$interval[which.max(steps_by_interval$steps)]
```

The peak is attained for the 835th interval.

```{r, echo=TRUE}
max(steps_by_interval$steps)
```

The peak is 206 steps.

#Imputing missing values

Total number of missing values in the dataset with NA:

```{r, echo=TRUE}
sum(is.na(activity))
```

As a strategy for filling in the missing values in the dataset, will be using the mean of the specific 5-minute interval for those observation that is missing.

```{r, echo=TRUE}
imputed_activitydata <- activity
for (i in 1:nrow(imputed_activitydata)) {
    if (is.na(imputed_activitydata$steps[i])) {
        interval_value <- imputed_activitydata$interval[i]
        steps_value <- steps_by_interval[
            steps_by_interval$interval == interval_value,]
        imputed_activitydata$steps[i] <- steps_value$steps
    }
}
```

```{r, echo=TRUE}
# calculate  total number of steps taken each day
imputed_steps_by_day <- aggregate(steps ~ date, imputed_activitydata, sum)
head(imputed_steps_by_day)
hist(imputed_steps_by_day$steps, main="Histogram of total number of steps per day (imputed)", xlab="Total number of steps in a day")

```
What is the impact of imputing missing data on the estimates of the total daily number of steps?

I would say there is minimum impact as  I used the average for a given interval, it pulled the averages closer to the inserted average value.


```{r, echo=TRUE}
mean(imputed_steps_by_day$steps)
median(imputed_steps_by_day$steps)
```

Mean remains the same and median values are higher after imputing missing data.


#Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE}

##Are there differences in activity patterns between weekdays and weekends?

imputed_activitydata['type_of_day'] <- weekdays(as.Date(imputed_activitydata$date))
imputed_activitydata$type_of_day[imputed_activitydata$type_of_day  %in% c('Saturday','Sunday') ] <- "weekend"
imputed_activitydata$type_of_day[imputed_activitydata$type_of_day != "weekend"] <- "weekday"

# convert type_of_day from character to factor
imputed_activitydata$type_of_day <- as.factor(imputed_activitydata$type_of_day)

# calculate average steps by interval across all days
imputed_steps_by_interval <- aggregate(steps ~ interval + type_of_day, imputed_activitydata, mean)

# creat a plot
qplot(interval, 
      steps, 
      data = imputed_steps_by_interval, 
      type = 'l', 
      geom=c("line"),
      xlab = "Interval", 
      ylab = "Number of steps", 
      main = "") +
    facet_wrap(~ type_of_day, ncol = 1)

```

From the graph, i guess there is more activities during week-end time.

